Lifting CDCL to Relational Domains

### 2016-10-15 domains.tex
more restructuring in 3

%-------------------------------------------------------------------------------
\paragraph{Safety equation}  
%
Based on above program representation, the safety equation solved by the ACDCL solver is the following:
%
\begin{equation}\label{eq:safety}
\varphi:= \bigwedge_{s\in\mathcal{P}} s \wedge \neg \bigwedge_{a\in\mathcal{A}} a
\text{ is unsatisfiable if and only if the program is safe.}
\end{equation}

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
The results that we present on lifting ACDL to relational domains hold
for any relational domain.
% 
In our prototype implementation and for illustrative purposes in this
paper we use a product domain $\mathcal{B}^b\times\mathcal{R}$ where
$\mathcal{B}$ is the Boolean domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$,
$b$ is the number of Boolean variables in the program, and $\mathcal{R}$
is a \emph{relational} domain over the numerical (bitvector) variables
in their respective data types (currently, signed and unsigned
integers are supported).
%
We instantiate $\mathcal{R}$ with the template polyhedra domain.

@@

@@

The meet irreducibles in the Boolean domain $\mathcal{B}$ are $\true$
and $\false$. The meet irreducibles in the interval domain are all
intervals that can be represented by a single inequality, i.e.\ $x \leq d$ or $x \geq d$ for constants $d$.

The meet irreducibles in the Boolean and template polyhedra domains
are all complementable.

@@

For example, the meet decomposition of the interval domain element
% $(x,y) \in [2,4]\times[3, 5]$ 
$2\leq x\leq 4 \wedge 3\leq y\leq 5$ is
the set $\{x\geq 2, x\leq 4, y\geq 3, y\leq 5\}$.
% \langle x:[2, 4] \rangle \in ItvDom$ which gives set of meet irreducibles, 
% $\{ \langle y \succeq 3 \rangle, \langle y \preceq 5 \rangle, 
% \langle x \succeq 2 \rangle, \langle x \preceq 4 \rangle \}$, that are 
% precisely complementable.

\subsection{Abstract Transformers}

In a typical abstract interpretation based Galois-connection setting
over an over-approximate domain, every concrete element has a unique
over-approximate representation in the abstract.  Likewise, every
concrete transformer is over-approximated by a unique abstract
transformer.  We now define an abstract deduction transformer.

@@

\subsection{Properties of Domain Elements for Learning}
%
\pscmt{TODO: Review by Peter}
An important property of a clause learning SAT solver 
is that each element of the partial assignments domain 
are decomposable into precisely complementable 
elements.  This property of the domain helps to guide 
the model search away from the conflicting region of the 
search space.

Most numerical abstract domains, such as intervals, octagons, 
polyhedra lacks precise complements, but they can be 
represented as intersections of complementable half-spaces, 
each of which have precise complements.  For example, intervals 
lack complements, but the decomposition of intervals into 
meet-irreducibles have complements.  
\[x \mapsto [2,5] \mathrel{\mathop{\longrightarrow}^{\mathrm{decompose}}} \{\langle x
\geq 2 \rangle \meet \langle x \leq 5 \rangle \} \]
Similarly, the complementation of the octagon in Figure~\ref{octagon} 
can be written as disjunction of:
\[(x<-2) \lor (x>1) \lor (y<-1) \lor (y>2) \lor (x+y>2) \lor (x-y>1) \lor (y-x>3) \lor (-x-y>2)\]

\Omit {
The commonly used library for numerical abstract domains  
is the APRON C library~\cite{apron}.  This library is 
used for the static analysis of the numerical variables 
of a program by abstract interpretation. APRON provides a 
C API interface to various abstract domains and libraries 
such as {\em BOX}, {\em OCTAGON}, {\em Convex Polyhedra} and
{\em Linear Equalities} library.  The aim of such analysis is 
to compute invariants over numerical variables in the 
program~\cite{se2011}. 
}
\Omit {
To this end, we implement our own template-based polyhedra domain and interval 
domain which supports complementation operator.  
%For example, the octagon in Figure~\ref{octagon} can be written as a conjunction of:
%\[(x>=-2) \land (x<=1) \land (y>=-1) \land (y<=2) \land (x+y<=2)
%\land (x-y<=1) \land (y-x<=3) \land (-x-y<=2)\] 
\pscmt{That's no valid motivation. We never complement a whole octagon, but just a
  meet-irreducible. It's trivial to do that with APRON. The reason was
a different one: APRON does not support all C operators, e.g. the bitwise
operators.} 
}

\subsection{Static Analysis Equations for Safety}
Static program analysis based on abstract interpretation~\cite{DBLP:conf/emsoft/Cousot07} 
perform safety analysis by computing fixed point to infer invariants 
over program variables.  However, bounded model checking (BMC) tries to search 
for a counterexample in a bounded execution trace, by symbolically executing a 
program up to a finite bound.  
Similar to BMC, ACDCL searches for a counterexample by solving the formula shown below.
For a set $N$ of program analysis constraints defined over a set 
of constraint variables $Var = \{X_i| i \in N\}$, representing a program
$\mathcal{P}$ and a concrete domain $\powerset(Env)$, the static analysis equation 
for safety of $\mathcal{P}$ is given as follows.  Here, $X_{init}, X_{Err}$ denote the initial valuation 
and the \rmcmt{final valuation} of constraint variables.
\[\varphi = X_{init} \subseteq Env \wedge \underset{p,q \in |\mathcal{P}|}
{\bigwedge} \{ X_p \subseteq post_{(p,q)}(X_q) \} \wedge X_{err} \supset \emptyset \]

>>>>

%-------------------------------------------------------------------------------
\paragraph{Safety problem}  
%
Based on above program representation, the safety problem solved by the ACDCL solver is the following:
%
\begin{equation}\label{eq:safety}
\varphi:= \bigwedge_{s\in\mathcal{P}} s \wedge \neg \bigwedge_{a\in\mathcal{A}} a
\text{ is unsatisfiable if and only if the program is safe.}
\end{equation}

% \subsection{Static Analysis Equations for Safety}
% Static program analysis based on abstract interpretation~\cite{DBLP:conf/emsoft/Cousot07} 
% perform safety analysis by computing fixed point to infer invariants 
% over program variables.  However, bounded model checking (BMC) tries to search 
% for a counterexample in a bounded execution trace, by symbolically executing a 
% program up to a finite bound.  
% Similar to BMC, ACDCL searches for a counterexample by solving the formula shown below.
% For a set $N$ of program analysis constraints defined over a set 
% of constraint variables $Var = \{X_i| i \in N\}$, representing a program
% $\mathcal{P}$ and a concrete domain $\powerset(Env)$, the static analysis equation 
% for safety of $\mathcal{P}$ is given as follows.  Here, $X_{init}, X_{Err}$ denote the initial valuation 
% and the \rmcmt{final valuation} of constraint variables.
% \[\varphi = X_{init} \subseteq Env \wedge \underset{p,q \in |\mathcal{P}|}
% {\bigwedge} \{ X_p \subseteq post_{(p,q)}(X_q) \} \wedge X_{err} \supset \emptyset \]

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
The results that we present on lifting ACDL to relational domains hold
for any relational domain.
% 
In our prototype implementation and for illustrative purposes in this
paper we use a product domain $\mathcal{B}^b\times\mathcal{R}$ where
$\mathcal{B}$ is the partial Boolean assignments domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$ (as
used in SAT solvers~\cite{sas12}), $b$ is the number of Boolean
variables in the program, and $\mathcal{R}$ is a \emph{relational}
domain over the numerical (bitvector) variables in their respective
data types (currently, signed and unsigned integers are supported).
%
We instantiate $\mathcal{R}$ with the template polyhedra domain.

@@

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains for Learning}
%
An important property of a clause learning SAT solver is that each
element of the partial Boolean assignments domain is decomposable into
complementable elements.  This property of the domain is required to
learn elements that help to guide the model search away from the
conflicting region of the search space.
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
complementable half-spaces, each of which has a complement. We
formalise this in the sequel.

@@

The meet irreducibles in the partial Boolean assignments domain
$\mathcal{B}$ are $\true$ and $\false$. The meet irreducibles in the
template polyhedra domain are all elements that concretise to
half-spaces, i.e.\ they can be represented by a single inequality. For
the interval domain, these are \ $x \leq d$ or $x \geq d$ for
constants $d$.

@@

For polyhedra this intuitively means that each polyhedron can be
written as an intersection of half-spaces.
%
For example, the meet decomposition of the interval domain element
% $(x,y) \in [2,4]\times[3, 5]$ 
$2\leq x\leq 4 \wedge 3\leq y\leq 5$ is
the set $\{x\geq 2, x\leq 4, y\geq 3, y\leq 5\}$.
% \langle x:[2, 4] \rangle \in ItvDom$ which gives set of meet irreducibles, 
% $\{ \langle y \succeq 3 \rangle, \langle y \preceq 5 \rangle, 
% \langle x \succeq 2 \rangle, \langle x \preceq 4 \rangle \}$, that are 
% precisely complementable.

The meet irreducibles in the partial Boolean assignments domain and
template polyhedra domains are all complementable.
%
Note that for template polyhedra, all non-meet-irreducible elements $e$
(except $\bot$ and $\top$) are not complementable within $A$.
%
However, we can complement each element in the meet decomposition of $e$ and
re-interpret the obtained set as a disjunction. 
%
Thus the complement of the octagon in Figure~\ref{octagon} 
can be written as a disjunction of meet irreducibles:
\[(x\leq -3) \lor (x\geq 2) \lor (y\leq -2) \lor (y\geq 3) \lor (x+y\geq 3) \lor (x-y\geq 2) \lor (y-x\geq 4) \lor (-x-y\geq 3)\]

%-------------------------------------------------------------------------------
\subsection{Abstract Transformers}
%
In a typical abstract interpretation based Galois-connection setting
over an over-approximate domain, every concrete element has a unique
over-approximate representation in the abstract.  Likewise, every
concrete transformer is over-approximated by a unique abstract
transformer.  We now define an abstract deduction transformer.

@@

% \Omit {
% The commonly used library for numerical abstract domains  
% is the APRON C library~\cite{apron}.  This library is 
% used for the static analysis of the numerical variables 
% of a program by abstract interpretation. APRON provides a 
% C API interface to various abstract domains and libraries 
% such as {\em BOX}, {\em OCTAGON}, {\em Convex Polyhedra} and
% {\em Linear Equalities} library.  The aim of such analysis is 
% to compute invariants over numerical variables in the 
% program~\cite{se2011}. 
% }
% \Omit {
% To this end, we implement our own template-based polyhedra domain and interval 
% domain which supports complementation operator.  
% %For example, the octagon in Figure~\ref{octagon} can be written as a conjunction of:
% %\[(x>=-2) \land (x<=1) \land (y>=-1) \land (y<=2) \land (x+y<=2)
% %\land (x-y<=1) \land (y-x<=3) \land (-x-y<=2)\] 
% \pscmt{That's no valid motivation. We never complement a whole octagon, but just a
%  meet-irreducible. It's trivial to do that with APRON. The reason was
% a different one: APRON does not support all C operators, e.g. the bitwise
% operators.} 
% }
